---
title: "Structural Equation Models"
---

For an overview of structural equation modeling, see my chapter on "[Structural Equation Modeling](https://isaactpetersen.github.io/Principles-Psychological-Assessment/structural-equation-modeling.html)" in the following textbook:

Petersen, I. T. (2025). *Principles of psychological assessment: With applied examples in R*. University of Iowa Libraries. <https://isaactpetersen.github.io/Principles-Psychological-Assessment>. <https://doi.org/10.25820/work.007199>

# Preamble

## Load Libraries

```{r}
library("lavaan")
library("semTools")
library("lcsm")
library("lavaanPlot")
library("lavaangui")
library("tidyverse")
```

## Load Data

```{r}
mydata_long <- read.csv("./data/mydata_long.csv")

longitudinalMI1 <- read.csv("./data/Bliese-Ployhart-2002-indicators-1.csv")
```

```{r}
#| eval: false

longitudinalMI2 <- read.csv("./data/Bliese-Ployhart-2002-indicators-2.csv")
longitudinalMI3 <- read.csv("./data/Bliese-Ployhart-2002-indicators-3.csv")
longitudinalMI4 <- read.csv("./data/Bliese-Ployhart-2002-indicators-4.csv")
```

# Pre-Model Computation

```{r}
mydata_long$sexFactor <- factor(mydata_long$sex)

mydata_long <- mydata_long |>
  group_by(ID) |>
  mutate(
    ses_personMean = mean(ses, na.rm = TRUE),
    ses_personMeanCentered = ses - ses_personMean,
    sleep_personMean = mean(sleep, na.rm = TRUE),
    sleep_personMeanCentered = sleep - sleep_personMean,
    depression_personMean = mean(depression, na.rm = TRUE),
    depression_personMeanCentered = depression - depression_personMean,
    anxiety_personMean = mean(anxiety, na.rm = TRUE),
    anxiety_personMeanCentered = anxiety - anxiety_personMean,
  ) |>
  ungroup()

mydata_wide <- mydata_long |>
  select(-age) |>
  tidyr::pivot_wider(
    names_from  = timepoint,
    values_from = c(ses, ses_personMeanCentered, sleep, sleep_personMeanCentered, depression, depression_personMeanCentered, anxiety, anxiety_personMeanCentered)
  )
```

# Latent Growth Curve Model

## Linear Growth Curve Model

### Model Syntax

In estimating linear latent growth curves of depression symptoms, this model specifies latent intercept and slope factors, `sex` as a time-invariant predictor of the latent intercept and slope, and `sleep` as a time-varying covariate predicting depression symptoms at each wave.

```{r}
linearGCM_syntax <- '
  # Intercept and slope
  intercept =~ 1*depression_1 + 1*depression_2 + 1*depression_3 + 1*depression_4
  slope =~ 0*depression_1 + 1*depression_2 + 2*depression_3 + 3*depression_4

  # Regression paths
  intercept ~ sexMale
  slope ~ sexMale
  
  # Time-varying covariates
  depression_1 ~ sleep_1
  depression_2 ~ sleep_2
  depression_3 ~ sleep_3
  depression_4 ~ sleep_4
  
  # Constrain observed intercepts to zero
  depression_1 ~ 0
  depression_2 ~ 0
  depression_3 ~ 0
  depression_4 ~ 0
  
  # Estimate mean of intercept and slope
  intercept ~ 1
  slope ~ 1
'
```

### Fit Model

```{r}
linearGCM_fit <- lavaan::sem(
  linearGCM_syntax,
  data = mydata_wide,
  missing = "ML", # FIML
  estimator = "MLR", # Huber-White robust standard errors (for heteroscedasticity and nonnormally distributed residuals)
  meanstructure = TRUE, # estimate means/intercepts
  fixed.x = FALSE)  # estimate (i.e., do not consider as fixed) means/variances/covariances of observed variables; allows missing information to be included as part of FIML
```

### Model Summary

```{r}
summary(
  linearGCM_fit,
  fit.measures = TRUE,
  standardized = TRUE,
  rsquare = TRUE)
```

### Modification Indices

```{r}
modificationindices(
  linearGCM_fit,
  sort. = TRUE)
```

### Plots

#### Path Diagram

```{r}
#| fig-cap: "Path Diagram"

lavaanPlot::lavaanPlot(
  linearGCM_fit,
  coefs = TRUE,
  #covs = TRUE,
  stand = TRUE)
```

```{r}
#| eval: false

lavaangui::plot_lavaan(linearGCM_fit)
```

## Quadratic Growth Curve Model

### Model Syntax

In estimating quadratic latent growth curves of depression symptoms, this model specifies latent intercept and latent linear and quadratic slope factors, `sex` as a time-invariant predictor of the latent intercept and slope, and `sleep` as a time-varying covariate predicting depression symptoms at each wave.

```{r}
quadraticGCM_syntax <- '
  # Intercept and slope
  intercept =~ 1*depression_1 + 1*depression_2 + 1*depression_3 + 1*depression_4
  linear =~ 0*depression_1 + 1*depression_2 + 2*depression_3 + 3*depression_4
  quadratic =~ 0*depression_1 + 1*depression_2 + 4*depression_3 + 9*depression_4

  # Regression paths
  intercept ~ sexMale
  linear ~ sexMale
  quadratic ~ sexMale
  
  # Time-varying covariates
  depression_1 ~ sleep_1
  depression_2 ~ sleep_2
  depression_3 ~ sleep_3
  depression_4 ~ sleep_4
  
  # Constrain observed intercepts to zero
  depression_1 ~ 0
  depression_2 ~ 0
  depression_3 ~ 0
  depression_4 ~ 0
  
  # Constrain residual variances to be equal (needed to address negative variance)
  depression_1 ~~ vardep*depression_1
  depression_2 ~~ vardep*depression_2
  depression_3 ~~ vardep*depression_3
  depression_4 ~~ vardep*depression_4
  
  # Estimate mean of intercept and slope
  intercept ~ 1
  linear ~ 1
  quadratic ~ 1
'
```

### Fit Model

```{r}
quadraticGCM_fit <- lavaan::sem(
  quadraticGCM_syntax,
  data = mydata_wide,
  missing = "ML",
  estimator = "MLR",
  meanstructure = TRUE,
  fixed.x = FALSE)
```

### Model Summary

```{r}
summary(
  quadraticGCM_fit,
  fit.measures = TRUE,
  standardized = TRUE,
  rsquare = TRUE)
```

### Modification Indices

```{r}
modificationindices(
  quadraticGCM_fit,
  sort. = TRUE)
```

### Plots

#### Path Diagram

```{r}
#| fig-cap: "Path Diagram"

lavaanPlot::lavaanPlot2(
  quadraticGCM_fit,
  stand = TRUE,
  coef_labels = TRUE)
```

```{r}
#| eval: false

lavaangui::plot_lavaan(quadraticGCM_fit)
```

## Latent Basis Growth Curve Model

### Model Syntax

In estimating latent basis growth curves of depression symptoms, this model specifies latent intercept and slope factors, with factor loadings for the slope freely estimated at timepoints 2 and 3 to capture nonlinear patterns of change, `sex` as a time-invariant predictor of the latent intercept and slope, and `sleep` as a time-varying covariate predicting depression symptoms at each wave.

```{r}
latentBasisGCM_syntax <- '
  # Intercept and slope
  intercept =~ 1*depression_1 + 1*depression_2 + 1*depression_3 + 1*depression_4
  slope =~ 0*depression_1 + a*depression_2 + b*depression_3 + 3*depression_4 # freely estimate the loadings for t2 and t3

  # Regression paths
  intercept ~ sexMale
  slope ~ sexMale
  
  # Time-varying covariates
  depression_1 ~ sleep_1
  depression_2 ~ sleep_2
  depression_3 ~ sleep_3
  depression_4 ~ sleep_4
  
  # Constrain observed intercepts to zero
  depression_1 ~ 0
  depression_2 ~ 0
  depression_3 ~ 0
  depression_4 ~ 0
  
  # Estimate mean of intercept and slope
  intercept ~ 1
  slope ~ 1
'
```

### Fit Model

```{r}
latentBasisGCM_fit <- lavaan::sem(
  latentBasisGCM_syntax,
  data = mydata_wide,
  missing = "ML",
  estimator = "MLR",
  meanstructure = TRUE,
  fixed.x = FALSE)
```

### Model Summary

```{r}
summary(
  latentBasisGCM_fit,
  fit.measures = TRUE,
  standardized = TRUE,
  rsquare = TRUE)
```

### Modification Indices

```{r}
modificationindices(
  latentBasisGCM_fit,
  sort. = TRUE)
```

### Plots

#### Path Diagram

```{r}
#| fig-cap: "Path Diagram"

lavaanPlot::lavaanPlot(
  latentBasisGCM_fit,
  coefs = TRUE,
  #covs = TRUE,
  stand = TRUE)
```

```{r}
#| eval: false

lavaangui::plot_lavaan(latentBasisGCM_fit)
```

# Compare Models

```{r}
anova(
  linearGCM_fit,
  quadraticGCM_fit
)

anova(
  linearGCM_fit,
  latentBasisGCM_fit
)

anova(
  latentBasisGCM_fit,
  quadraticGCM_fit
)
```

# Latent Change Score Model

## Model Syntax

This model examines reciprocal (i.e., bidirectional), dynamic (i.e., changing) relations between latent change in depression and latent change in anxiety.
Each construct is modeled with a latent intercept (constant change factor), a proportional change factor (latent true score predicting its own change), and autoregressive effects of prior change scores.
The cross-domain coupling parameters evaluate whether latent change in depression predicts subsequent latent change in anxiety, and vice versa.

```{r}
bivariateLCSM_syntax <- lcsm::specify_bi_lcsm(
  timepoints = 4,
  var_x = "depression_",
  model_x = list(
    alpha_constant = TRUE, # alpha = intercept (constant change factor)
    beta = TRUE, # beta = proportional change factor (latent true score predicting its change score)
    phi = TRUE), # phi = autoregression of change scores
  var_y = "anxiety_",
  model_y = list(
    alpha_constant = TRUE, # alpha = intercept (constant change factor)
    beta = TRUE, # beta = proportional change factor (latent true score predicting its change score)
    phi = TRUE), # phi = autoregression of change scores
  coupling = list(
    delta_lag_xy = TRUE,
    delta_lag_yx = TRUE),
  change_letter_x = "g",
  change_letter_y = "j")

cat(bivariateLCSM_syntax)
```

## Fit Model

```{r}
bivariateLCSM_fit <- lcsm::fit_bi_lcsm(
  data = mydata_wide,
  var_x = c("depression_1","depression_2","depression_3","depression_4"),
  var_y = c("anxiety_1","anxiety_2","anxiety_3","anxiety_4"),
  model_x = list(
    alpha_constant = TRUE, # alpha = intercept (constant change factor)
    beta = TRUE, # beta = proportional change factor (latent true score predicting its change score)
    phi = TRUE), # phi = autoregression of change scores
  model_y = list(
    alpha_constant = TRUE, # alpha = intercept (constant change factor)
    beta = TRUE, # beta = proportional change factor (latent true score predicting its change score)
    phi = TRUE), # phi = autoregression of change scores
  coupling = list(
    delta_lag_xy = TRUE,
    xi_lag_yx = TRUE),
  fixed.x = FALSE
  )
```

## Model Summary

```{r}
summary(
  bivariateLCSM_fit,
  fit.measures = TRUE,
  standardized = TRUE,
  rsquare = TRUE)

lcsm::extract_param(bivariateLCSM_fit)
```

## Modification Indices

```{r}
modificationindices(
  bivariateLCSM_fit,
  sort. = TRUE)
```

## Plots

### Path Diagram

```{r}
#| fig-cap: "Path Diagram"

lavaanPlot::lavaanPlot2(
  bivariateLCSM_fit,
  stand = TRUE,
  coef_labels = TRUE)
```

```{r}
#| eval: false

lavaangui::plot_lavaan(bivariateLCSM_fit)
```

```{r}
#| eval: false

lcsm::plot_lcsm(
  lavaan_object = bivariateLCSM_fit,
  lcsm = "bivariate",
  lavaan_syntax = bivariateLCSM_syntax,
  edge.label.cex = .9,
  lcsm_colours = TRUE)
```

To plot model-implied trajectories, see this resource:

<https://thechangelab.stanford.edu/tutorials/growth-modeling/latent-change-score-modeling-lavaan/> (archived at <https://perma.cc/A2KZ-QPD9>)

# Cross-Lagged Panel Model

## Model Syntax

This cross-lagged panel model examines whether depression predicts later anxiety (i.e., cross-lagged paths), controlling for prior levels of anxiety (i.e., autoregressive effects), and simultaneously examining the reverse.
Concurrent covariances between depression and anxiety at each timepoint account for shared variance at the same occasion.

```{r}
clpm_syntax <- '
  # Autoregressive Paths
  depression_4 ~ depression_3
  depression_3 ~ depression_2
  depression_2 ~ depression_1
  
  anxiety_4 ~ anxiety_3
  anxiety_3 ~ anxiety_2
  anxiety_2 ~ anxiety_1
  
  # Concurrent Covariances
  depression_1 ~~ anxiety_1
  depression_2 ~~ anxiety_2
  depression_3 ~~ anxiety_3
  depression_4 ~~ anxiety_4
  
  # Cross-Lagged Paths
  depression_4 ~ anxiety_3
  depression_3 ~ anxiety_2
  depression_2 ~ anxiety_1
  
  anxiety_4 ~ depression_3
  anxiety_3 ~ depression_2
  anxiety_2 ~ depression_1
'
```

## Fit Model

```{r}
clpm_fit <- lavaan::sem(
  clpm_syntax,
  data = mydata_wide,
  missing = "ML",
  estimator = "MLR",
  meanstructure = TRUE,
  fixed.x = FALSE)
```

## Model Summary

```{r}
summary(
  clpm_fit,
  fit.measures = TRUE,
  standardized = TRUE,
  rsquare = TRUE)
```

## Modification Indices

```{r}
modificationindices(
  clpm_fit,
  sort. = TRUE)
```

## Plots

### Path Diagram

```{r}
#| fig-cap: "Path Diagram"

lavaanPlot::lavaanPlot(
  clpm_fit,
  coefs = TRUE,
  #covs = TRUE,
  stand = TRUE)
```

```{r}
#| eval: false

lavaangui::plot_lavaan(clpm_fit)
```

# Random-Intercepts Cross-Lagged Panel Model

Adapted from:

- Mulder & Hamaker (2021): <https://doi.org/10.1080/10705511.2020.1784738>
  - <https://jeroendmulder.github.io/RI-CLPM/lavaan.html> (archived at <https://perma.cc/2K6A-WUJQ>)
- Mund & Nestler (2017): <https://osf.io/a4dhk>

## Model Syntax

This random-intercepts cross-lagged panel model examines whether depression predicts later anxiety (i.e., cross-lagged paths), controlling for prior levels of anxiety (i.e., autoregressive effects) and controlling for a person's general latent level of anxiety, and simultaneously examining the reverse.
Concurrent covariances between depression and anxiety at each timepoint account for shared variance at the same occasion.

```{r}
riclpm_syntax <- '
  # Random Intercepts
  depression =~ 1*depression_1 + 1*depression_2 + 1*depression_3 + 1*depression_4
  anxiety =~ 1*anxiety_1 + 1*anxiety_2 + 1*anxiety_3 + 1*anxiety_4
  
  # Create Within-Person Centered Variables
  wdep1 =~ 1*depression_1
  wdep2 =~ 1*depression_2
  wdep3 =~ 1*depression_3
  wdep4 =~ 1*depression_4
  
  wanx1 =~ 1*anxiety_1
  wanx2 =~ 1*anxiety_2
  wanx3 =~ 1*anxiety_3
  wanx4 =~ 1*anxiety_4
  
  # Autoregressive Paths
  wdep4 ~ wdep3
  wdep3 ~ wdep2
  wdep2 ~ wdep1
  
  wanx4 ~ wanx3
  wanx3 ~ wanx2
  wanx2 ~ wanx1
  
  # Concurrent Covariances
  wdep1 ~~ wanx1
  wdep2 ~~ wanx2
  wdep3 ~~ wanx3
  wdep4 ~~ wanx4
  
  # Cross-Lagged Paths
  wdep4 ~ wanx3
  wdep3 ~ wanx2
  wdep2 ~ wanx1
  
  wanx4 ~ wdep3
  wanx3 ~ wdep2
  wanx2 ~ wdep1
  
  # Variance and Covariance of Random Intercepts
  depression ~~ depression
  anxiety ~~ anxiety
  depression ~~ anxiety
  
  # Variances of Within-Person Centered Variables
  wdep1 ~~ wdep1
  wdep2 ~~ wdep2
  wdep3 ~~ wdep3
  wdep4 ~~ wdep4
  
  wanx1 ~~ wanx1
  wanx2 ~~ wanx2
  wanx3 ~~ wanx3
  wanx4 ~~ wanx4
  
  # Fix Error Variances of Observed Variables to Zero
  depression_1 ~~ 0*depression_1
  depression_2 ~~ 0*depression_2
  depression_3 ~~ 0*depression_3
  depression_4 ~~ 0*depression_4
  
  anxiety_1 ~~ 0*anxiety_1
  anxiety_2 ~~ 0*anxiety_2
  anxiety_3 ~~ 0*anxiety_3
  anxiety_4 ~~ 0*anxiety_4
  
  # Fix the Covariances Between the Random Intercepts and the Latents at T1 to Zero
  wdep1 ~~ 0*depression
  wdep1 ~~ 0*anxiety
  
  wanx1 ~~ 0*depression
  wanx1 ~~ 0*anxiety
  
  # Estimate Observed Intercepts
  depression_1 ~ 1
  depression_2 ~ 1
  depression_3 ~ 1
  depression_4 ~ 1
  
  anxiety_1 ~ 1
  anxiety_2 ~ 1
  anxiety_3 ~ 1
  anxiety_4 ~ 1
  
  # Fix the Means of the Latents to Zero
  wdep1 ~ 0*1
  wdep2 ~ 0*1
  wdep3 ~ 0*1
  wdep4 ~ 0*1
  
  wanx1 ~ 0*1
  wanx2 ~ 0*1
  wanx3 ~ 0*1
  wanx4 ~ 0*1
  
  depression ~ 0*1
  anxiety ~ 0*1
'
```

## Fit Model

```{r}
riclpm_fit <- lavaan::sem(
  riclpm_syntax,
  data = mydata_wide,
  missing = "ML",
  estimator = "MLR",
  fixed.x = FALSE,
  bounds = "standard"
)
```

Modify the syntax to address the model fit errors.

## Model Summary

```{r}
#| eval: false

summary(
  withinPersonCLPM_fit,
  fit.measures = TRUE,
  standardized = TRUE,
  rsquare = TRUE)
```

## Modification Indices

```{r}
#| eval: false

modificationindices(
  riclpm_fit,
  sort. = TRUE)
```

## Path Diagram

```{r}
#| eval: false

lavaanPlot::lavaanPlot(
  riclpm_fit,
  coefs = TRUE,
  #covs = TRUE,
  stand = TRUE)
```

```{r}
#| eval: false

lavaangui::plot_lavaan(riclpm_fit)
```

# Multilevel SEM

## Model Syntax

This multilevel structural equation model disaggregates within- and between-person effects to examine associations with depression.
At the within-person level, deviations in sleep, anxiety, and SES from a person's usual levels predict concurrent deviations in depression.
At the between-person level, person-level averages across time of sleep, anxiety, and SES, along with sex, predict overall differences in depression across individuals.

```{r}
multilevelSEM_syntax <- '
    level: 1
        depression ~ sleep_personMeanCentered + anxiety_personMeanCentered + ses_personMeanCentered
    level: 2
        depression ~ sleep_personMean + anxiety_personMean + ses_personMean + sexMale
'
```

## Fit Model

```{r}
multilevelSEM_fit <- lavaan::sem(
  multilevelSEM_syntax,
  data = mydata_long,
  cluster = "ID",
  #missing = "ML", # commenting out due to error
  estimator = "MLR",
  fixed.x = FALSE)
```

## Model Summary

```{r}
summary(
  multilevelSEM_fit,
  fit.measures = TRUE,
  standardized = TRUE,
  rsquare = TRUE)
```

## Modification Indices

```{r}
modificationindices(
  multilevelSEM_fit,
  sort. = TRUE)
```

## Path Diagram

```{r}
lavaanPlot::lavaanPlot(
  multilevelSEM_fit,
  coefs = TRUE,
  #covs = TRUE,
  stand = TRUE)
```

```{r}
#| eval: false

lavaangui::plot_lavaan(multilevelSEM_fit)
```

# Longitudinal Measurement Invariance

We will follow the approach suggested by Widaman et al. (2010).

Widaman, K. F., Ferrer, E., & Conger, R. D. (2010). Factorial invariance within longitudinal structural equation models: Measuring the same construct across time. *Child Development Perspectives*, *4*(1), 10–18. <https://doi.org/10.1111/j.1750-8606.2009.00110.x>

For the longitudinal measurement invariance models, we will use simulated data from <https://mycourses.aalto.fi/mod/assign/view.php?id=1203047> (archived at <https://perma.cc/QTL2-ZHX2>).
As noted [here](https://mycourses.aalto.fi/mod/assign/view.php?id=1203047) (archived at <https://perma.cc/QTL2-ZHX2>), "The data are in wide format where the variables are coded as [VARIABLE NAME][time index][indicator index]."

Questions:

1. Which level of longitudinal measurement invariance holds for the `Bliese-Ployhard-2000-indicators-1.csv` dataset (saved in object `longitudinalMI1`)?
1. Repeat the tests of longitudinal measurement invariance for the remaining datasets: `Bliese-Ployhard-2000-indicators-2.csv`, `Bliese-Ployhard-2000-indicators-3.csv`, and `Bliese-Ployhard-2000-indicators-4.csv`.
Which level of longitudinal measurement invariance holds for each dataset?

The answers to these questions are [here](answers.qmd).

## Configural Invariance

A configural invariance model whether there are the same number of latent factors across time and whether the indicators load onto the same latent factor(s) across time.

1. Standardize the latent factor(s) at T1 (i.e., fix the mean to zero and the variance to one)
1. For each latent construct, constrain the first indicator's factor loading to be the same across time
1. For each latent construct, constrain the first indicator's intercept to be the same across time

### Model Syntax

```{r}
configuralInvariance_syntax <- '
  # Factor Loadings
  jobsat_1 =~ NA*loadj1*JOBSAT11 + JOBSAT12 + JOBSAT13
  jobsat_2 =~ NA*loadj1*JOBSAT21 + JOBSAT22 + JOBSAT23
  jobsat_3 =~ NA*loadj1*JOBSAT31 + JOBSAT32 + JOBSAT33
  
  ready_1 =~ NA*loadr1*READY11 + READY12 + READY13
  ready_2 =~ NA*loadr1*READY21 + READY22 + READY23
  ready_3 =~ NA*loadr1*READY31 + READY32 + READY33
  
  commit_1 =~ NA*loadc1*COMMIT11 + COMMIT12 + COMMIT13
  commit_2 =~ NA*loadc1*COMMIT21 + COMMIT22 + COMMIT23
  commit_3 =~ NA*loadc1*COMMIT31 + COMMIT32 + COMMIT33
  
  # Factor Identification: Standardize Factors at T1
  
  ## Fix Factor Means at T1 to Zero
  jobsat_1 ~ 0*1
  ready_1 ~ 0*1
  commit_1 ~ 0*1
  
  ## Fix Factor Variances at T1 to One
  jobsat_1 ~~ 1*jobsat_1
  ready_1 ~~ 1*ready_1
  commit_1 ~~ 1*commit_1
  
  # Freely Estimate Factor Means at T2 and T3 (relative to T1)
  jobsat_2 ~ 1
  jobsat_3 ~ 1
  
  ready_2 ~ 1
  ready_3 ~ 1
  
  commit_2 ~ 1
  commit_3 ~ 1
  
  # Freely Estimate Factor Variances at T2 and T3 (relative to T1)
  jobsat_2 ~~ jobsat_2
  jobsat_3 ~~ jobsat_3
  
  ready_2 ~~ ready_2
  ready_3 ~~ ready_3
  
  commit_2 ~~ commit_2
  commit_3 ~~ commit_3
  
  # Fix Intercepts of Indicator 1 Across Time
  JOBSAT11 ~ intj1*1
  JOBSAT21 ~ intj1*1
  JOBSAT31 ~ intj1*1
  
  READY11 ~ intr1*1
  READY21 ~ intr1*1
  READY31 ~ intr1*1
  
  COMMIT11 ~ intc1*1
  COMMIT21 ~ intc1*1
  COMMIT31 ~ intc1*1
  
  # Free Intercepts of Remaining Manifest Variables
  JOBSAT12 ~ 1
  JOBSAT13 ~ 1
  JOBSAT22 ~ 1
  JOBSAT23 ~ 1
  JOBSAT32 ~ 1
  JOBSAT33 ~ 1
  
  READY12 ~ 1
  READY13 ~ 1
  READY22 ~ 1
  READY23 ~ 1
  READY32 ~ 1
  READY33 ~ 1
  
  COMMIT12 ~ 1
  COMMIT13 ~ 1
  COMMIT22 ~ 1
  COMMIT23 ~ 1
  COMMIT32 ~ 1
  COMMIT33 ~ 1
  
  # Estimate Residual Variances of Manifest Variables
  JOBSAT11 ~~ JOBSAT11
  JOBSAT12 ~~ JOBSAT12
  JOBSAT13 ~~ JOBSAT13
  JOBSAT21 ~~ JOBSAT21
  JOBSAT22 ~~ JOBSAT22
  JOBSAT23 ~~ JOBSAT23
  JOBSAT31 ~~ JOBSAT31
  JOBSAT32 ~~ JOBSAT32
  JOBSAT33 ~~ JOBSAT33
  
  READY11 ~~ READY11
  READY12 ~~ READY12
  READY13 ~~ READY13
  READY21 ~~ READY21
  READY22 ~~ READY22
  READY23 ~~ READY23
  READY31 ~~ READY31
  READY32 ~~ READY32
  READY33 ~~ READY33
  
  COMMIT11 ~~ COMMIT11
  COMMIT12 ~~ COMMIT12
  COMMIT13 ~~ COMMIT13
  COMMIT21 ~~ COMMIT21
  COMMIT22 ~~ COMMIT22
  COMMIT23 ~~ COMMIT23
  COMMIT31 ~~ COMMIT31
  COMMIT32 ~~ COMMIT32
  COMMIT33 ~~ COMMIT33
'
```

### Fit Model

```{r}
configuralInvariance_fit <- cfa(
  configuralInvariance_syntax,
  data = longitudinalMI1,
  missing = "ML",
  estimator = "MLR",
  meanstructure = TRUE,
  #std.lv = TRUE,
  fixed.x = FALSE)
```

### Model Summary

```{r}
summary(
  configuralInvariance_fit,
  fit.measures = TRUE,
  standardized = TRUE,
  rsquare = TRUE)
```

### Modification Indices

```{r}
modificationindices(
  configuralInvariance_fit,
  sort. = TRUE)
```

### Path Diagram

```{r}
#| fig-cap: "Path Diagram"

lavaanPlot::lavaanPlot2(
  configuralInvariance_fit,
  stand = TRUE,
  coef_labels = TRUE)
```

```{r}
#| eval: false

lavaangui::plot_lavaan(configuralInvariance_fit)
```

## Configural Invariance With Correlated Residuals Within-Indicator Across Time

1. Standardize the latent factor(s) at T1 (i.e., fix the mean to zero and the variance to one)
1. For each latent construct, constrain the first indicator's factor loading to be the same across time
1. For each latent construct, constrain the first indicator's intercept to be the same across time
1. **Allow within-indicator residuals to covary across time**

### Model Syntax

```{r}
configuralInvarianceCorrelatedResiduals_syntax <- '
  # Factor Loadings
  jobsat_1 =~ NA*loadj1*JOBSAT11 + JOBSAT12 + JOBSAT13
  jobsat_2 =~ NA*loadj1*JOBSAT21 + JOBSAT22 + JOBSAT23
  jobsat_3 =~ NA*loadj1*JOBSAT31 + JOBSAT32 + JOBSAT33
  
  ready_1 =~ NA*loadr1*READY11 + READY12 + READY13
  ready_2 =~ NA*loadr1*READY21 + READY22 + READY23
  ready_3 =~ NA*loadr1*READY31 + READY32 + READY33
  
  commit_1 =~ NA*loadc1*COMMIT11 + COMMIT12 + COMMIT13
  commit_2 =~ NA*loadc1*COMMIT21 + COMMIT22 + COMMIT23
  commit_3 =~ NA*loadc1*COMMIT31 + COMMIT32 + COMMIT33
  
  # Factor Identification: Standardize Factors at T1
  
  ## Fix Factor Means at T1 to Zero
  jobsat_1 ~ 0*1
  ready_1 ~ 0*1
  commit_1 ~ 0*1
  
  ## Fix Factor Variances at T1 to One
  jobsat_1 ~~ 1*jobsat_1
  ready_1 ~~ 1*ready_1
  commit_1 ~~ 1*commit_1
  
  # Freely Estimate Factor Means at T2 and T3 (relative to T1)
  jobsat_2 ~ 1
  jobsat_3 ~ 1
  
  ready_2 ~ 1
  ready_3 ~ 1
  
  commit_2 ~ 1
  commit_3 ~ 1
  
  # Freely Estimate Factor Variances at T2 and T3 (relative to T1)
  jobsat_2 ~~ jobsat_2
  jobsat_3 ~~ jobsat_3
  
  ready_2 ~~ ready_2
  ready_3 ~~ ready_3
  
  commit_2 ~~ commit_2
  commit_3 ~~ commit_3
  
  # Fix Intercepts of Indicator 1 Across Time
  JOBSAT11 ~ intj1*1
  JOBSAT21 ~ intj1*1
  JOBSAT31 ~ intj1*1
  
  READY11 ~ intr1*1
  READY21 ~ intr1*1
  READY31 ~ intr1*1
  
  COMMIT11 ~ intc1*1
  COMMIT21 ~ intc1*1
  COMMIT31 ~ intc1*1
  
  # Free Intercepts of Remaining Manifest Variables
  JOBSAT12 ~ 1
  JOBSAT13 ~ 1
  JOBSAT22 ~ 1
  JOBSAT23 ~ 1
  JOBSAT32 ~ 1
  JOBSAT33 ~ 1
  
  READY12 ~ 1
  READY13 ~ 1
  READY22 ~ 1
  READY23 ~ 1
  READY32 ~ 1
  READY33 ~ 1
  
  COMMIT12 ~ 1
  COMMIT13 ~ 1
  COMMIT22 ~ 1
  COMMIT23 ~ 1
  COMMIT32 ~ 1
  COMMIT33 ~ 1
  
  # Estimate Residual Variances of Manifest Variables
  JOBSAT11 ~~ JOBSAT11
  JOBSAT12 ~~ JOBSAT12
  JOBSAT13 ~~ JOBSAT13
  JOBSAT21 ~~ JOBSAT21
  JOBSAT22 ~~ JOBSAT22
  JOBSAT23 ~~ JOBSAT23
  JOBSAT31 ~~ JOBSAT31
  JOBSAT32 ~~ JOBSAT32
  JOBSAT33 ~~ JOBSAT33
  
  READY11 ~~ READY11
  READY12 ~~ READY12
  READY13 ~~ READY13
  READY21 ~~ READY21
  READY22 ~~ READY22
  READY23 ~~ READY23
  READY31 ~~ READY31
  READY32 ~~ READY32
  READY33 ~~ READY33
  
  COMMIT11 ~~ COMMIT11
  COMMIT12 ~~ COMMIT12
  COMMIT13 ~~ COMMIT13
  COMMIT21 ~~ COMMIT21
  COMMIT22 ~~ COMMIT22
  COMMIT23 ~~ COMMIT23
  COMMIT31 ~~ COMMIT31
  COMMIT32 ~~ COMMIT32
  COMMIT33 ~~ COMMIT33
  
  # Residual Covariances Within Indicator Across Time
  JOBSAT11 ~~ JOBSAT21
  JOBSAT21 ~~ JOBSAT31
  JOBSAT11 ~~ JOBSAT31
  
  JOBSAT12 ~~ JOBSAT22
  JOBSAT22 ~~ JOBSAT32
  JOBSAT12 ~~ JOBSAT32
  
  JOBSAT13 ~~ JOBSAT23
  JOBSAT23 ~~ JOBSAT33
  JOBSAT13 ~~ JOBSAT33
  
  READY11 ~~ READY21
  READY21 ~~ READY31
  READY11 ~~ READY31
  
  READY12 ~~ READY22
  READY22 ~~ READY32
  READY12 ~~ READY32
  
  READY13 ~~ READY23
  READY23 ~~ READY33
  READY13 ~~ READY33
  
  COMMIT11 ~~ COMMIT21
  COMMIT21 ~~ COMMIT31
  COMMIT11 ~~ COMMIT31
  
  COMMIT12 ~~ COMMIT22
  COMMIT22 ~~ COMMIT32
  COMMIT12 ~~ COMMIT32
  
  COMMIT13 ~~ COMMIT23
  COMMIT23 ~~ COMMIT33
  COMMIT13 ~~ COMMIT33
'
```

### Fit Model

```{r}
configuralInvarianceCorrelatedResiduals_fit <- cfa(
  configuralInvarianceCorrelatedResiduals_syntax,
  data = longitudinalMI1,
  missing = "ML",
  estimator = "MLR",
  meanstructure = TRUE,
  #std.lv = TRUE,
  fixed.x = FALSE)
```

### Model Summary

```{r}
summary(
  configuralInvarianceCorrelatedResiduals_fit,
  fit.measures = TRUE,
  standardized = TRUE,
  rsquare = TRUE)
```

### Modification Indices

```{r}
modificationindices(
  configuralInvarianceCorrelatedResiduals_fit,
  sort. = TRUE)
```

### Path Diagram

```{r}
#| fig-cap: "Path Diagram"

lavaanPlot::lavaanPlot2(
  configuralInvarianceCorrelatedResiduals_fit,
  stand = TRUE,
  coef_labels = TRUE)
```

```{r}
#| eval: false

lavaangui::plot_lavaan(configuralInvarianceCorrelatedResiduals_fit)
```

### Compare Models

```{r}
anova(
  configuralInvariance_fit,
  configuralInvarianceCorrelatedResiduals_fit
)
```

## Metric ("Weak Factorial") Invariance

A metric invariance model evaluates whether the items' factor loadings are the same across time.

1. Standardize the latent factor(s) at T1 (i.e., fix the mean to zero and the variance to one)
1. For each latent construct, constrain the first indicator's intercept to be the same across time
1. Allow within-indicator residuals to covary across time
1. **For each indicator, constrain its factor loading to be the same across time**

### Model Syntax

```{r}
metricInvariance_syntax <- '
  # Factor Loadings
  jobsat_1 =~ NA*loadj1*JOBSAT11 + loadj2*JOBSAT12 + loadj3*JOBSAT13
  jobsat_2 =~ NA*loadj1*JOBSAT21 + loadj2*JOBSAT22 + loadj3*JOBSAT23
  jobsat_3 =~ NA*loadj1*JOBSAT31 + loadj2*JOBSAT32 + loadj3*JOBSAT33
  
  ready_1 =~ NA*loadr1*READY11 + loadr2*READY12 + loadr3*READY13
  ready_2 =~ NA*loadr1*READY21 + loadr2*READY22 + loadr3*READY23
  ready_3 =~ NA*loadr1*READY31 + loadr2*READY32 + loadr3*READY33
  
  commit_1 =~ NA*loadc1*COMMIT11 + loadc2*COMMIT12 + loadc3*COMMIT13
  commit_2 =~ NA*loadc1*COMMIT21 + loadc2*COMMIT22 + loadc3*COMMIT23
  commit_3 =~ NA*loadc1*COMMIT31 + loadc2*COMMIT32 + loadc3*COMMIT33
  
  # Factor Identification: Standardize Factors at T1
  
  ## Fix Factor Means at T1 to Zero
  jobsat_1 ~ 0*1
  ready_1 ~ 0*1
  commit_1 ~ 0*1
  
  ## Fix Factor Variances at T1 to One
  jobsat_1 ~~ 1*jobsat_1
  ready_1 ~~ 1*ready_1
  commit_1 ~~ 1*commit_1
  
  # Freely Estimate Factor Means at T2 and T3 (relative to T1)
  jobsat_2 ~ 1
  jobsat_3 ~ 1
  
  ready_2 ~ 1
  ready_3 ~ 1
  
  commit_2 ~ 1
  commit_3 ~ 1
  
  # Freely Estimate Factor Variances at T2 and T3 (relative to T1)
  jobsat_2 ~~ jobsat_2
  jobsat_3 ~~ jobsat_3
  
  ready_2 ~~ ready_2
  ready_3 ~~ ready_3
  
  commit_2 ~~ commit_2
  commit_3 ~~ commit_3
  
  # Fix Intercepts of Indicator 1 Across Time
  JOBSAT11 ~ intj1*1
  JOBSAT21 ~ intj1*1
  JOBSAT31 ~ intj1*1
  
  READY11 ~ intr1*1
  READY21 ~ intr1*1
  READY31 ~ intr1*1
  
  COMMIT11 ~ intc1*1
  COMMIT21 ~ intc1*1
  COMMIT31 ~ intc1*1
  
  # Free Intercepts of Remaining Manifest Variables
  JOBSAT12 ~ 1
  JOBSAT13 ~ 1
  JOBSAT22 ~ 1
  JOBSAT23 ~ 1
  JOBSAT32 ~ 1
  JOBSAT33 ~ 1
  
  READY12 ~ 1
  READY13 ~ 1
  READY22 ~ 1
  READY23 ~ 1
  READY32 ~ 1
  READY33 ~ 1
  
  COMMIT12 ~ 1
  COMMIT13 ~ 1
  COMMIT22 ~ 1
  COMMIT23 ~ 1
  COMMIT32 ~ 1
  COMMIT33 ~ 1
  
  # Estimate Residual Variances of Manifest Variables
  JOBSAT11 ~~ JOBSAT11
  JOBSAT12 ~~ JOBSAT12
  JOBSAT13 ~~ JOBSAT13
  JOBSAT21 ~~ JOBSAT21
  JOBSAT22 ~~ JOBSAT22
  JOBSAT23 ~~ JOBSAT23
  JOBSAT31 ~~ JOBSAT31
  JOBSAT32 ~~ JOBSAT32
  JOBSAT33 ~~ JOBSAT33
  
  READY11 ~~ READY11
  READY12 ~~ READY12
  READY13 ~~ READY13
  READY21 ~~ READY21
  READY22 ~~ READY22
  READY23 ~~ READY23
  READY31 ~~ READY31
  READY32 ~~ READY32
  READY33 ~~ READY33
  
  COMMIT11 ~~ COMMIT11
  COMMIT12 ~~ COMMIT12
  COMMIT13 ~~ COMMIT13
  COMMIT21 ~~ COMMIT21
  COMMIT22 ~~ COMMIT22
  COMMIT23 ~~ COMMIT23
  COMMIT31 ~~ COMMIT31
  COMMIT32 ~~ COMMIT32
  COMMIT33 ~~ COMMIT33
  
  # Residual Covariances Within Indicator Across Time
  JOBSAT11 ~~ JOBSAT21
  JOBSAT21 ~~ JOBSAT31
  JOBSAT11 ~~ JOBSAT31
  
  JOBSAT12 ~~ JOBSAT22
  JOBSAT22 ~~ JOBSAT32
  JOBSAT12 ~~ JOBSAT32
  
  JOBSAT13 ~~ JOBSAT23
  JOBSAT23 ~~ JOBSAT33
  JOBSAT13 ~~ JOBSAT33
  
  READY11 ~~ READY21
  READY21 ~~ READY31
  READY11 ~~ READY31
  
  READY12 ~~ READY22
  READY22 ~~ READY32
  READY12 ~~ READY32
  
  READY13 ~~ READY23
  READY23 ~~ READY33
  READY13 ~~ READY33
  
  COMMIT11 ~~ COMMIT21
  COMMIT21 ~~ COMMIT31
  COMMIT11 ~~ COMMIT31
  
  COMMIT12 ~~ COMMIT22
  COMMIT22 ~~ COMMIT32
  COMMIT12 ~~ COMMIT32
  
  COMMIT13 ~~ COMMIT23
  COMMIT23 ~~ COMMIT33
  COMMIT13 ~~ COMMIT33
'
```

### Fit Model

```{r}
metricInvariance_fit <- cfa(
  metricInvariance_syntax,
  data = longitudinalMI1,
  missing = "ML",
  estimator = "MLR",
  meanstructure = TRUE,
  #std.lv = TRUE,
  fixed.x = FALSE)
```

### Model Summary

```{r}
summary(
  metricInvariance_fit,
  fit.measures = TRUE,
  standardized = TRUE,
  rsquare = TRUE)
```

### Modification Indices

```{r}
modificationindices(
  metricInvariance_fit,
  sort. = TRUE)
```

### Path Diagram

```{r}
#| fig-cap: "Path Diagram"

lavaanPlot::lavaanPlot2(
  metricInvariance_fit,
  stand = TRUE,
  coef_labels = TRUE)
```

```{r}
#| eval: false

lavaangui::plot_lavaan(metricInvariance_fit)
```

### Compare Models

```{r}
anova(
  configuralInvarianceCorrelatedResiduals_fit,
  metricInvariance_fit
)
```

## Scalar ("Strong Factorial") Invariance

A scalar invariance model evaluates whether the items' intercepts are the same across time.

1. Standardize the latent factor(s) at T1 (i.e., fix the mean to zero and the variance to one)
1. Allow within-indicator residuals to covary across time
1. For each indicator, constrain its factor loading to be the same across time
1. **For each indicator, constrain its intercept to be the same across time**

### Model Syntax

```{r}
scalarInvariance_syntax <- '
  # Factor Loadings
  jobsat_1 =~ NA*loadj1*JOBSAT11 + loadj2*JOBSAT12 + loadj3*JOBSAT13
  jobsat_2 =~ NA*loadj1*JOBSAT21 + loadj2*JOBSAT22 + loadj3*JOBSAT23
  jobsat_3 =~ NA*loadj1*JOBSAT31 + loadj2*JOBSAT32 + loadj3*JOBSAT33
  
  ready_1 =~ NA*loadr1*READY11 + loadr2*READY12 + loadr3*READY13
  ready_2 =~ NA*loadr1*READY21 + loadr2*READY22 + loadr3*READY23
  ready_3 =~ NA*loadr1*READY31 + loadr2*READY32 + loadr3*READY33
  
  commit_1 =~ NA*loadc1*COMMIT11 + loadc2*COMMIT12 + loadc3*COMMIT13
  commit_2 =~ NA*loadc1*COMMIT21 + loadc2*COMMIT22 + loadc3*COMMIT23
  commit_3 =~ NA*loadc1*COMMIT31 + loadc2*COMMIT32 + loadc3*COMMIT33
  
  # Factor Identification: Standardize Factors at T1
  
  ## Fix Factor Means at T1 to Zero
  jobsat_1 ~ 0*1
  ready_1 ~ 0*1
  commit_1 ~ 0*1
  
  ## Fix Factor Variances at T1 to One
  jobsat_1 ~~ 1*jobsat_1
  ready_1 ~~ 1*ready_1
  commit_1 ~~ 1*commit_1
  
  # Freely Estimate Factor Means at T2 and T3 (relative to T1)
  jobsat_2 ~ 1
  jobsat_3 ~ 1
  
  ready_2 ~ 1
  ready_3 ~ 1
  
  commit_2 ~ 1
  commit_3 ~ 1
  
  # Freely Estimate Factor Variances at T2 and T3 (relative to T1)
  jobsat_2 ~~ jobsat_2
  jobsat_3 ~~ jobsat_3
  
  ready_2 ~~ ready_2
  ready_3 ~~ ready_3
  
  commit_2 ~~ commit_2
  commit_3 ~~ commit_3
  
  # Fix Intercepts of Indicators Across Time
  JOBSAT11 ~ intj1*1
  JOBSAT21 ~ intj1*1
  JOBSAT31 ~ intj1*1
  JOBSAT12 ~ intj2*1
  JOBSAT22 ~ intj2*1
  JOBSAT32 ~ intj2*1
  JOBSAT13 ~ intj3*1
  JOBSAT23 ~ intj3*1
  JOBSAT33 ~ intj3*1
  
  READY11 ~ intr1*1
  READY21 ~ intr1*1
  READY31 ~ intr1*1
  READY12 ~ intr2*1
  READY22 ~ intr2*1
  READY32 ~ intr2*1
  READY13 ~ intr3*1
  READY23 ~ intr3*1
  READY33 ~ intr3*1
  
  COMMIT11 ~ intc1*1
  COMMIT21 ~ intc1*1
  COMMIT31 ~ intc1*1
  COMMIT12 ~ intc2*1
  COMMIT22 ~ intc2*1
  COMMIT32 ~ intc2*1
  COMMIT13 ~ intc3*1
  COMMIT23 ~ intc3*1
  COMMIT33 ~ intc3*1
  
  # Estimate Residual Variances of Manifest Variables
  JOBSAT11 ~~ JOBSAT11
  JOBSAT12 ~~ JOBSAT12
  JOBSAT13 ~~ JOBSAT13
  JOBSAT21 ~~ JOBSAT21
  JOBSAT22 ~~ JOBSAT22
  JOBSAT23 ~~ JOBSAT23
  JOBSAT31 ~~ JOBSAT31
  JOBSAT32 ~~ JOBSAT32
  JOBSAT33 ~~ JOBSAT33
  
  READY11 ~~ READY11
  READY12 ~~ READY12
  READY13 ~~ READY13
  READY21 ~~ READY21
  READY22 ~~ READY22
  READY23 ~~ READY23
  READY31 ~~ READY31
  READY32 ~~ READY32
  READY33 ~~ READY33
  
  COMMIT11 ~~ COMMIT11
  COMMIT12 ~~ COMMIT12
  COMMIT13 ~~ COMMIT13
  COMMIT21 ~~ COMMIT21
  COMMIT22 ~~ COMMIT22
  COMMIT23 ~~ COMMIT23
  COMMIT31 ~~ COMMIT31
  COMMIT32 ~~ COMMIT32
  COMMIT33 ~~ COMMIT33
  
  # Residual Covariances Within Indicator Across Time
  JOBSAT11 ~~ JOBSAT21
  JOBSAT21 ~~ JOBSAT31
  JOBSAT11 ~~ JOBSAT31
  
  JOBSAT12 ~~ JOBSAT22
  JOBSAT22 ~~ JOBSAT32
  JOBSAT12 ~~ JOBSAT32
  
  JOBSAT13 ~~ JOBSAT23
  JOBSAT23 ~~ JOBSAT33
  JOBSAT13 ~~ JOBSAT33
  
  READY11 ~~ READY21
  READY21 ~~ READY31
  READY11 ~~ READY31
  
  READY12 ~~ READY22
  READY22 ~~ READY32
  READY12 ~~ READY32
  
  READY13 ~~ READY23
  READY23 ~~ READY33
  READY13 ~~ READY33
  
  COMMIT11 ~~ COMMIT21
  COMMIT21 ~~ COMMIT31
  COMMIT11 ~~ COMMIT31
  
  COMMIT12 ~~ COMMIT22
  COMMIT22 ~~ COMMIT32
  COMMIT12 ~~ COMMIT32
  
  COMMIT13 ~~ COMMIT23
  COMMIT23 ~~ COMMIT33
  COMMIT13 ~~ COMMIT33
'
```

### Fit Model

```{r}
scalarInvariance_fit <- cfa(
  scalarInvariance_syntax,
  data = longitudinalMI1,
  missing = "ML",
  estimator = "MLR",
  meanstructure = TRUE,
  #std.lv = TRUE,
  fixed.x = FALSE)
```

### Model Summary

```{r}
summary(
  scalarInvariance_fit,
  fit.measures = TRUE,
  standardized = TRUE,
  rsquare = TRUE)
```

### Modification Indices

```{r}
modificationindices(
  scalarInvariance_fit,
  sort. = TRUE)
```

### Path Diagram

```{r}
#| fig-cap: "Path Diagram"

lavaanPlot::lavaanPlot2(
  scalarInvariance_fit,
  stand = TRUE,
  coef_labels = TRUE)
```

```{r}
#| eval: false

lavaangui::plot_lavaan(scalarInvariance_fit)
```

### Compare Models

```{r}
anova(
  metricInvariance_fit,
  scalarInvariance_fit
)
```

## Residual ("Strict Factorial") Invariance

A residual invariance model evaluates whether the items' residual variances are the same across time.

1. Standardize the latent factor(s) at T1 (i.e., fix the mean to zero and the variance to one)
1. Allow within-indicator residuals to covary across time
1. For each indicator, constrain its factor loading to be the same across time
1. For each indicator, constrain its intercept to be the same across time
1. **For each indicator, constrain its residual variance to be the same across time**

### Model Syntax

```{r}
residualInvariance_syntax <- '
  # Factor Loadings
  jobsat_1 =~ NA*loadj1*JOBSAT11 + loadj2*JOBSAT12 + loadj3*JOBSAT13
  jobsat_2 =~ NA*loadj1*JOBSAT21 + loadj2*JOBSAT22 + loadj3*JOBSAT23
  jobsat_3 =~ NA*loadj1*JOBSAT31 + loadj2*JOBSAT32 + loadj3*JOBSAT33
  
  ready_1 =~ NA*loadr1*READY11 + loadr2*READY12 + loadr3*READY13
  ready_2 =~ NA*loadr1*READY21 + loadr2*READY22 + loadr3*READY23
  ready_3 =~ NA*loadr1*READY31 + loadr2*READY32 + loadr3*READY33
  
  commit_1 =~ NA*loadc1*COMMIT11 + loadc2*COMMIT12 + loadc3*COMMIT13
  commit_2 =~ NA*loadc1*COMMIT21 + loadc2*COMMIT22 + loadc3*COMMIT23
  commit_3 =~ NA*loadc1*COMMIT31 + loadc2*COMMIT32 + loadc3*COMMIT33
  
  # Factor Identification: Standardize Factors at T1
  
  ## Fix Factor Means at T1 to Zero
  jobsat_1 ~ 0*1
  ready_1 ~ 0*1
  commit_1 ~ 0*1
  
  ## Fix Factor Variances at T1 to One
  jobsat_1 ~~ 1*jobsat_1
  ready_1 ~~ 1*ready_1
  commit_1 ~~ 1*commit_1
  
  # Freely Estimate Factor Means at T2 and T3 (relative to T1)
  jobsat_2 ~ 1
  jobsat_3 ~ 1
  
  ready_2 ~ 1
  ready_3 ~ 1
  
  commit_2 ~ 1
  commit_3 ~ 1
  
  # Freely Estimate Factor Variances at T2 and T3 (relative to T1)
  jobsat_2 ~~ jobsat_2
  jobsat_3 ~~ jobsat_3
  
  ready_2 ~~ ready_2
  ready_3 ~~ ready_3
  
  commit_2 ~~ commit_2
  commit_3 ~~ commit_3
  
  # Constrain Intercepts of Indicators Across Time
  JOBSAT11 ~ intj1*1
  JOBSAT21 ~ intj1*1
  JOBSAT31 ~ intj1*1
  JOBSAT12 ~ intj2*1
  JOBSAT22 ~ intj2*1
  JOBSAT32 ~ intj2*1
  JOBSAT13 ~ intj3*1
  JOBSAT23 ~ intj3*1
  JOBSAT33 ~ intj3*1
  
  READY11 ~ intr1*1
  READY21 ~ intr1*1
  READY31 ~ intr1*1
  READY12 ~ intr2*1
  READY22 ~ intr2*1
  READY32 ~ intr2*1
  READY13 ~ intr3*1
  READY23 ~ intr3*1
  READY33 ~ intr3*1
  
  COMMIT11 ~ intc1*1
  COMMIT21 ~ intc1*1
  COMMIT31 ~ intc1*1
  COMMIT12 ~ intc2*1
  COMMIT22 ~ intc2*1
  COMMIT32 ~ intc2*1
  COMMIT13 ~ intc3*1
  COMMIT23 ~ intc3*1
  COMMIT33 ~ intc3*1
  
  # Constrain Residual Variances of Indicators Across Time
  JOBSAT11 ~~ resj1*JOBSAT11
  JOBSAT21 ~~ resj1*JOBSAT21
  JOBSAT31 ~~ resj1*JOBSAT31
  JOBSAT12 ~~ resj2*JOBSAT12
  JOBSAT22 ~~ resj2*JOBSAT22
  JOBSAT32 ~~ resj2*JOBSAT32
  JOBSAT13 ~~ resj3*JOBSAT13
  JOBSAT23 ~~ resj3*JOBSAT23
  JOBSAT33 ~~ resj3*JOBSAT33
  
  READY11 ~~ resr1*READY11
  READY21 ~~ resr1*READY21
  READY31 ~~ resr1*READY31
  READY12 ~~ resr2*READY12
  READY22 ~~ resr2*READY22
  READY32 ~~ resr2*READY32
  READY13 ~~ resr3*READY13
  READY23 ~~ resr3*READY23
  READY33 ~~ resr3*READY33
  
  COMMIT11 ~~ resc1*COMMIT11
  COMMIT21 ~~ resc1*COMMIT21
  COMMIT31 ~~ resc1*COMMIT31
  COMMIT12 ~~ resc2*COMMIT12
  COMMIT22 ~~ resc2*COMMIT22
  COMMIT32 ~~ resc2*COMMIT32
  COMMIT13 ~~ resc3*COMMIT13
  COMMIT23 ~~ resc3*COMMIT23
  COMMIT33 ~~ resc3*COMMIT33
  
  # Residual Covariances Within Indicator Across Time
  JOBSAT11 ~~ JOBSAT21
  JOBSAT21 ~~ JOBSAT31
  JOBSAT11 ~~ JOBSAT31
  
  JOBSAT12 ~~ JOBSAT22
  JOBSAT22 ~~ JOBSAT32
  JOBSAT12 ~~ JOBSAT32
  
  JOBSAT13 ~~ JOBSAT23
  JOBSAT23 ~~ JOBSAT33
  JOBSAT13 ~~ JOBSAT33
  
  READY11 ~~ READY21
  READY21 ~~ READY31
  READY11 ~~ READY31
  
  READY12 ~~ READY22
  READY22 ~~ READY32
  READY12 ~~ READY32
  
  READY13 ~~ READY23
  READY23 ~~ READY33
  READY13 ~~ READY33
  
  COMMIT11 ~~ COMMIT21
  COMMIT21 ~~ COMMIT31
  COMMIT11 ~~ COMMIT31
  
  COMMIT12 ~~ COMMIT22
  COMMIT22 ~~ COMMIT32
  COMMIT12 ~~ COMMIT32
  
  COMMIT13 ~~ COMMIT23
  COMMIT23 ~~ COMMIT33
  COMMIT13 ~~ COMMIT33
'
```

### Fit Model

```{r}
residualInvariance_fit <- cfa(
  residualInvariance_syntax,
  data = longitudinalMI1,
  missing = "ML",
  estimator = "MLR",
  meanstructure = TRUE,
  #std.lv = TRUE,
  fixed.x = FALSE)
```

### Model Summary

```{r}
summary(
  residualInvariance_fit,
  fit.measures = TRUE,
  standardized = TRUE,
  rsquare = TRUE)
```

### Modification Indices

```{r}
modificationindices(
  residualInvariance_fit,
  sort. = TRUE)
```

### Path Diagram

```{r}
#| fig-cap: "Path Diagram"

lavaanPlot::lavaanPlot2(
  residualInvariance_fit,
  stand = TRUE,
  coef_labels = TRUE)
```

```{r}
#| eval: false

lavaangui::plot_lavaan(residualInvariance_fit)
```

### Compare Models

```{r}
anova(
  scalarInvariance_fit,
  residualInvariance_fit
)

round(cbind(
  configural = inspect(configuralInvarianceCorrelatedResiduals_fit, "fit.measures"), 
  weak = inspect(metricInvariance_fit, "fit.measures"),
  strong = inspect(scalarInvariance_fit, "fit.measures"),
  strict = inspect(residualInvariance_fit, "fit.measures")), 3)
```
